Case study validations of automatic bullet matching

### 2018-10-11 index.Rmd
addressing first couple of edits

# Introduction and Background
In current practice, firearms and toolmark examiners (FTE) evaluate the similarity of striae on bullets by placing the evidence together with a bullet from the suspected gun under a comparison microscope and visually classify similarity according to the theory of firearms identification \citep{identification} as one of identification, inconclusive or exclusion. Exact guidelines for this classification vary from lab to lab, e.g.\ some labs will exclude only on the basis of non-matching class characteristics (such as direction of the twist in rifling, land length or number of lands, or type of rifling). In some labs,  CMS (consecutively matching striae) as defined by \citet{biasotti} \svp{are} used as a way to quantify the strength of a match.

## 3D Imaging Technology
\svp{3D characterization of bullet surfaces has been considered since 1958 \citeauthor{davis1968introduction}, but the technology was not sufficiently advanced for analysis based on 3D measurements to be a viable option. Approaches to utilizing 3D measurment data were proposed starting in 1999 \citeauthor{dekinderAutomatedComparisonsBullet1999}, and have been further developed by \citet{bachrachDevelopment3DbasedAutomated2002,xieAutomatedBulletidentificationSystem2009, chuPilotStudyAutomated2010}. These approaches utilize 3D surface measurements directly to characterize the topology of bullet surfaces, rather than using visual or image comparison techniques. In many cases, they also provide some level of automation of the comparison process, with the goal of reducing human biases by augmenting the visual information with 3D measurements. This allows for examination not only of the peaks in the bullet signature, but also for comparison of the valleys, which are more difficult to assess visually except in relation to the peaks. }

\svp{Commonly, approaches derived from 3D surface measurements use some features which are developed to mimic the perception of FTEs\citep{luAutomatedBulletIdentification2014}. Class characteristics, which are shared by a group of firearms with the same rifling design, manufacturer, and tooling process, are typically evaluated first, as a mismatch on class charcteristics is sufficient to conclude that two samples do not match. Automated approaches to estimation of width of the land engraved area and twist angle were described in \citet{chuPilotStudyAutomated2010}. Individual characteristics, which are not shared by all members of a class, can also be automatically assessed from 3D surface measurements; in some cases, these features mimic features used by FTEs during image analysis. One of the most common features used to assess the strength of a match based on individual characteristics is the cross-correlation function (ccf), which is described in several studies \citep{maNISTBulletSignature2004, vorburgerApplicationsCrosscorrelationFunctions2011, chuPilotStudyAutomated2010}. Additional features proposed for automatically assessing match strength also include signature distance \citep{maNISTBulletSignature2004}, striae depth and width \citep{ComputerIdentificationBullets1978}, and consecutive matching striae \citep{chuAutomaticIdentificationBullet2013}.}

In the \svp{matching algorithm proposed in \citet{aoas}, many of the features described above are combined to produce a similarity score} based on 3d topographic scans of land engraved areas (LEAs). From each land engraved area a signature is derived by

@@

Quantitative features describing the similarity of two signature are extracted. Features include characteristics of the signature that firearms and toolmark examiners base their visual assessment on, such as \svp{striae depth}, number of matching striae, and the number of consecutively matching striae (CMS). We also include statistical features, such as cross-correlation and Euclidean distance between the signatures.

@@

The algorithm in \citet{aoas} is trained on scans from Hamby sets 252 and 173 made available through \citet{nist}. Set 173 was originally published as Hamby set 44\svp{, but was mislabeled.} \svp{The scans used to fit the model were taken at a resolution of \svp{Probably need to mention that the NIST scans have a different resolution, which may affect how the results generalize?}

Here, we are considering three validation sets: 
\begin{itemize}
\item {\bf Hamby set 44} 
Hamby set 44 is one set of the Hamby study \citep{hamby}. Each Hamby set  consists of a total of 35 bullets fired through ten consecutively manufactured barrels of Rugers P85. Each set consists of 20 known bullets (two from each of the ten barrels) and 15 questioned bullets of unknown origin. Note that all Hamby sets are closed sets\svp{; that is,} all questioned bullets are fired through one of the ten barrels. The ammunition used for this set were 9 mm Luger 115 Grain Full Metal Jacket  from the Winchester Ammunition Company.
\item {\bf Phoenix PD}
Tylor Klep from Phoenix PD provided sets of known test fires and questioned bullets: the set of known bullets consists of three test fires (B1, B2, B3) from each of eight different barrels (A9, C8, F6, L5, M2, P7, R3, U10). Ten questioned bullets were provided (B, E, H, J, K N, Q, T, Y, Z). This set is an open set\svp{; that is,}\ it is not known in advance whether all (or any) of the questioned bullets are fired from the known barrels.

Land engraved areas for each of the six lands of each bullet were scanned by Bill Henderson (Sensofar).
\item {\bf Houston FSI}
This study was set up by Melissa Nally and Kasi Kirksey from FSI Houston. Three test sets based on ten consecutively rifled Ruger LCP barrels (A, B, C, D, E, F, G, H, I, J) and three other, non-consecutively rifled Ruger LCP barrels (R1, R2, R3). Each test set consists of three test fires each from five consecutively rifled barrels. Additionally, ten questioned bullets are provided for each kit. The ammunition used in both test fires and the questioned bullets were Remington UMC 9mm Luger Full Metal Jackets. All three of the test sets are open\svp{; that is,}\ not every one of the questioned bullets is fired from the five known barrels in each of the test set.
\end{itemize}

Scans of all land engraved areas were taken on a Sensofar Confocal Light microscope at 20x magnification resulting \svp{in} a resolution of 0.645 microns per pixel. If not indicated otherwise, scans were taken at the Roy J Carver high resolution microscopy lab at Iowa State University.

Case studies were chosen such that difficulty for the matching algorithm increases \svp{with each case study}: Hamby set 44 is part of the Hamby study. The algorithm in \citet{aoas} was trained on sets 173 and 252, so the bullets in Hamby 44 are of the same type of ammunition and are fired through the same barrels as the bullets in the training set.

\svp{The }Phoenix PD set \svp{uses} Ruger XXX barrels, \svp{which }are different from the barrels in the \svp{Hamby sets used for }training\svp{ the algorithm}, but the barrels are \svp{rifled similarly to} the Ruger P-85 barrels.

\svp{The} Houston sets\svp{ use} Ruger LCP barrels. These barrels are first rifled traditionally, i.e. similar to the Ruger P-85, but are double fired. \svp{\{I still don't understand what double fired means in this context\}} The second firing happens after the rifling. This second firing has the potential to introduce some subclass characteristics, i.e. anomalies that are shared between similarly \svp{rifled? manufactured?} barrels. This should make the automatic matching, and in particular, the classification as different-source land engraved areas/bullets, harder.

>>>>

# Introduction and Background
In current practice, firearms and toolmark examiners (FTE) evaluate the similarity of striae on bullets by placing the evidence together with a bullet from the suspected gun under a comparison microscope and visually classify similarity according to the theory of firearms identification \citep{identification} as one of identification, inconclusive or exclusion. Exact guidelines for this classification vary from lab to lab, e.g.\ some labs will exclude only on the basis of non-matching class characteristics (such as direction of the twist in rifling, land length or number of lands, or type of rifling). In some labs,  CMS (consecutively matching striae) as defined by \citet{biasotti} is used as a measure to quantify the strength of a match.

## 3D Imaging Technology
3D characterization of bullet surfaces has been considered since 1958 \citeauthor{davis1968introduction}, but the technology was not sufficiently advanced for analysis based on 3D measurements to be a viable option. Approaches to utilizing 3D measurement data were proposed starting in 1999 \citeauthor{dekinderAutomatedComparisonsBullet1999}, and have been further developed by \citet{bachrachDevelopment3DbasedAutomated2002,xieAutomatedBulletidentificationSystem2009, chuPilotStudyAutomated2010}. These approaches utilize 3D surface measurements directly to characterize the topology of bullet surfaces, rather than using visual or image comparison techniques. In many cases, they also provide some level of automation of the comparison process, with the goal of reducing human biases by augmenting the visual information with 3D measurements. This allows for examination not only of the peaks in the bullet signature, but also for comparison of the valleys, which are more difficult to assess visually except in relation to the peaks.

Commonly, approaches derived from 3D surface measurements use some features which are developed to mimic the perception of FTEs\citep{luAutomatedBulletIdentification2014}. Class characteristics, which are shared by a group of firearms with the same rifling design, manufacturer, and tooling process, are typically evaluated first, as a mismatch on class charcteristics is sufficient for an exclusion. Automated approaches to estimation of width of the land engraved area and twist angle were described in \citet{chuPilotStudyAutomated2010}. Individual characteristics, which are not shared by all members of a class, can also be automatically assessed from 3D surface measurements; in some cases, these features mimic features used by FTEs during image analysis. One of the most common features used to assess the strength of a match based on individual characteristics is the cross-correlation function (ccf), which is described in several studies \citep{maNISTBulletSignature2004, vorburgerApplicationsCrosscorrelationFunctions2011, chuPilotStudyAutomated2010}. Additional features proposed for automatically assessing match strength also include signature distance \citep{maNISTBulletSignature2004}, striae depth and width \citep{ComputerIdentificationBullets1978}, and consecutive matching striae \citep{chuAutomaticIdentificationBullet2013}.

In the matching algorithm proposed in \citet{aoas}, several of the features described above are combined to produce a similarity score based on 3d topographic scans of land engraved areas (LEAs). From each land engraved area a signature is derived by

@@

Quantitative features describing the similarity of two signature are extracted. Features include characteristics of the signature that firearms and toolmark examiners base their visual assessment on, such as striae depth, number of matching striae, and the number of consecutively matching striae (CMS). We also include statistical features, such as cross-correlation and Euclidean distance between the signatures.

@@

The algorithm in \citet{aoas} is trained on scans from Hamby sets 252 and 173 made available through \citet{nist}. Set 173 was originally published as Hamby set 44. This mislabelling has been correct now. The scans used to train the model were taken at 20 fold magnification for a resolution of 1.5625 microns per pixel. \hh{While magnification is generally of interest in microscopy, {\it resolution} -- usually measured in microns per pixel -- is of more interest for 3D topographic measurements, as it determines the operative level of available data.}

\hh{XXX We need a gentler intro to the validation sets - maybe some discussion of training and test sets?}

Here, we are considering three validation sets: 
\begin{itemize}
\item {\bf Hamby set 44} 
Hamby set 44 is one set of the Hamby study \citep{hamby}. Each Hamby set  consists of a total of 35 bullets fired through ten consecutively manufactured barrels of Rugers P85. Each set consists of 20 known bullets (two from each of the ten barrels) and 15 questioned bullets of unknown origin. Note that all Hamby sets are closed sets; that is, all questioned bullets are fired through one of the ten barrels. The ammunition used for this set were 9 mm Luger 115 Grain Full Metal Jacket  from the Winchester Ammunition Company.
\item {\bf Phoenix PD}
Tylor Klep from Phoenix PD provided sets of known test fires and questioned bullets: the set of known bullets consists of three test fires (B1, B2, B3) from each of eight different barrels (A9, C8, F6, L5, M2, P7, R3, U10). Ten questioned bullets were provided (B, E, H, J, K N, Q, T, Y, Z). This set is an open set; that is,\ it is not known in advance whether all (or any) of the questioned bullets are fired from the known barrels.

Land engraved areas for each of the six lands of each bullet were scanned by Bill Henderson (Sensofar).
\item {\bf Houston FSI}
This study was set up by Melissa Nally and Kasi Kirksey from FSI Houston. Three test sets based on ten consecutively rifled Ruger LCP barrels (A, B, C, D, E, F, G, H, I, J) and three other, non-consecutively rifled Ruger LCP barrels (R1, R2, R3). Each test set consists of three test fires each from five consecutively rifled barrels. Additionally, ten questioned bullets are provided for each kit. The ammunition used in both test fires and the questioned bullets were Remington UMC 9mm Luger Full Metal Jackets. All three of the test sets are open; that is,\ not every one of the questioned bullets is fired from the five known barrels in each of the test set.
\end{itemize}

Scans of all land engraved areas were taken on a Sensofar Confocal Light microscope at 20x magnification resulting in a resolution of 0.645 microns per pixel. If not indicated otherwise, scans were taken at the Roy J Carver high resolution microscopy lab at Iowa State University.

Case studies were chosen such that difficulty for the matching algorithm increases with each case study: Hamby set 44 is part of the Hamby study. The algorithm in \citet{aoas} was trained on sets 173 and 252, so the bullets in Hamby 44 are of the same type of ammunition and are fired through the same barrels as the bullets in the training set.

The Phoenix PD set uses Ruger XXX barrels, which are different from the barrels in the Hamby sets used for training the algorithm, but the barrels are rifled similarly to the Ruger P-85 barrels.

\svp{The} Houston sets\svp{ use} Ruger LCP barrels. These barrels are first rifled traditionally, i.e. similar to the Ruger P-85, but are treated to a secondary round of heating after barreling. This second round of heat exposure has the potential to introduce some subclass characteristics, i.e. anomalies that are shared between different barrels manufactured in this manner. This should make the automatic matching harder, and in particular, the classification as different-source land engraved areas/bullets.
